---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE, results='hide', message=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
require(binnr)
data(titanic)
levels(titanic$Embarked)[levels(titanic$Embarked) == ""] <- "C"
```

# What is binnr?

`binnr` is an R package that helps modelers build scorecards. Scorecard models
need to be more than predictive. Regulatory oversight often necessitates they be
transparent as well. Model transparency is difficult to enforce in more
predictive, non-linear methods such as neural networks, random forest, or
gradient boosted decision trees. Often direct modeler intervention is required
to ensure proper treatment is given for values of independent variables.

`binnr` attempts to solve these problems by providing interactive variable
manipulation facilities giving the modeler total control over how variables
are treated in scorecard models.

## Installing binnr

The easiest way to install `binnr` is by using the `devtools` package:

```{r, eval=FALSE}
if (!require(devtools)) install.packages("devtools")
devtools::install_github(repo="Zelazny7/binnr")
```

## Overview of binnr

`binnr` is comprised of three main modeling steps: Bin, Fit, & Adjust. Each of
these steps are outlined in the following sections.

### Bin

The Bin phase prepares a data set of explanatory features for modeling by
discretizing variables and wrapping in an object that provides data manipulation
capabilities. How variables are treated by the binning step depends on their
type. Numeric features such as age or income are discretized using a performance
metric. For example, if the performance variable is binary, the numeric
independent variables are discretized using information value.

If the variable is a categorical field (factors in R) no discretization can be 
performed; however, the features is still wrapped by a bin object and can be 
interacted with as such.

```{r, echo=TRUE, eval=FALSE}
mod <- bin(data=titanic[,-1], y=titanic$Survived)
```

Calling the `bin` function returns a Scorecard object. The Scorecard object
contains methods for manipulating and fitting the discretized variables, or
"bins" in binnr parlance. The Scorecard object maintains a list of bins as and
keeps track of the operations requested by the modelers. It also contains a list
of fitted scorecard models that can be adjusted or reviewed. These capabilities
will be expanded upon in later sections.

```{r, echo=FALSE, eval=TRUE, results='hide'}
mod <- bin(data=titanic[,-1], y=titanic$Survived)
```

```{r, echo=TRUE, prompt=TRUE}
mod
```

Bin variables can be accessed directly. Though as we'll see later, this is not
recommended. But for illustration purposes, the following output is the kind of
information a bin object typically stores:

```{r, echo=TRUE, prompt=TRUE}
mod$variables$Pclass
```

Additionally, bin objects can be plotted by calling their `plot` method. 

```{r, echo=TRUE, prompt=TRUE,fig.height=5, fig.width=7, fig.align='center'}
mod$variables$Pclass$plot()
```

### Fit

The second step of scorecard developement using `binnr` is fitting a model.
`binnr` uses the `glmnet` package under the hood to fit a regularized
regression model. The discretized variables have their observed performance
values substituted for their actual input values to create a completely 
continuous dataset.

> For example, if an age variable is discretized using binary performance the 
> resulting substitution uses the observed weight-of-evidence rather than the 
> age itself.

Performance substitution is used to put all predictors on the same scale and 
force a linear relationship between each predictor and the response variables.

```{r, echo=TRUE, eval=TRUE, results='hide', collapse=TRUE}
mod$fit("model 1", "initial model with all variables")
```
```{r, echo=TRUE, prompt=TRUE, collapse=TRUE}
mod
```

Once a model is fit, it can be predicted as well. This returns a score value
appropriate for the type of model that was fit to the response variable. In the
case of binary performance, a binomial response was selected resulting in a 
logit.

```{r, echo=TRUE, eval=FALSE}
pred <- mod$predict()
```
```{r, echo=FALSE, eval=TRUE, results='hide'}
pred <- mod$predict()
```
```{r, echo=TRUE, prompt=TRUE}
head(pred)
```

### Adjust

The most important functionality provided by `binnr` is the manipulation of 
bin objects.

Bins can be collapsed:

```{r, echo=TRUE, eval=TRUE}
mod$variables$Pclass$collapse(c(1,3))
mod$variables$Pclass$show()

# Alternate syntax
# mod$variables$Pclass - c(1,3)
```

Bins can be expanded:

```{r, echo=TRUE, eval=TRUE}
mod$variables$Pclass$expand(1)
mod$variables$Pclass$show()

# Alternate syntax
# mod$variables$Pclass + 1
```


